Effects of pulsed hollow electron lens operation on the beam core in HL-LHC: \\First experimental studies and simulations.

### 2018-01-30 resex.tex
complete first revision

In a recent review, the need for such an active halo control system for HL-LHC has been assessed with the conclusion that it would considerably increase the margin and reduce the risk for machine protection compared to the current LHC collimation system being a passive halo control system  \cite{helreview}. In view of the general need of active halo control for future high power accelerators and HL-LHC in particular, different methods have been studied in the last recent years \cite{helreview_bruce}, of which the HEL is considered the superior device at least in case of the HL-LHC \cite{helreview} and is also considered for other future high energy colliders like HE-LHC and FCC-hh \cite{helhcparam2011,fcc_coll_ipac2017}.

The beneficial effect of an active halo control system is however abrogated if also the beam core is affected as this would ultimately lead to a degradation of the performance due to losses of particles in the core region and emittance growth. In this paper, we will concentrate on this aspect in view of future HEL to be installed in the HL-LHC. We will summarize possible sources of perturbations of the beam core concentrating in particular on the case of pulsed operation and with the focus on the beam experiments at the LHC performed in this context.

This paper is structured as follows: Section~\ref{sec:hel} gives an introduction to the concept of HELs and summarizes the design parameters of the HL-LHC HEL. Section~\ref{sec:core} is dedicated to describing the sources of a residual field from the HEL in the core region. Section~\ref{sec:exp} describes then in detail the two beam experiments conducted at the LHC to study the effect on the beam core in case of a pulsed HEL operation. This includes the detailed analysis of the observed losses, emittance and transverse beam distribution changes. To the knowledge of the authors, the observed distribution changes presented in this paper have never been measured before in the context of a resonant excitation. A resonant excitation has been previously studied at the Tevatron in the context of the HEL experiments \cite{hel_tevatron_stancari} and the abort gap cleaning used in operation \cite{hel_tevatron_abortgap_zhang}. Both studies however only concentrated on the losses and emittance changes without measuring the detailed changes of the distribution itself. In addition, the presented experiments also provide scalings of the losses and emittance growth with the excitation amplitude. This allows to provide first tolerances for the residual field of the HL-LHC HEL at the beam core and also to compare the obtained results with simulations and thus benchmark the same.

\section{Hollow electron lens for HL-LHC\label{sec:hel}}
\subsection{General Overview\label{sec:hel:intro}}
In general, electron lenses are DC or pulsed low energy electron beams. The electron beam is generated in an electron gun, then guided and confined by strong solenoids and finally dumped onto a collector. Exemplary for the conceptual design of all electron lenses, the HL-LHC HEL is shown in Fig.~\ref{fig:hel_layout}.
\begin{figure}[h]
	\includegraphics[width=1.0\linewidth]{hel_layout_epbeam}% Here is how to import EPS art
	\caption{\label{fig:hel_layout} Layout of HL-LHC HEL.}
\end{figure}
The circulating beam, in case of the LHC the proton beam, is then affected by the electromagnetic field of the electron beam. For the application of active halo control, the electron beam needs to generate an electromagnetic field at the location of the halo particles while leaving the core untouched. This can be achieved by using a uniform hollow distribution in radius $r=\sqrt{x^2+y^2}$ with inner radius $R_1$ and outer radius $R_2$. In this case, the circulating proton beam experiences a radial kick $\theta(r)$
\begin{equation}\label{eq:field_1}
\theta(r)=\frac{f(r)}{(r/R_2)}\cdot \theta_{\rm max},
\end{equation}
where $f(r)$ is a shape function with
\begin{equation}\label{eq:field_2}
f(r) =
\begin{cases} 0 &,\quad r< R_1,\\
\frac{r^2-R_1^2}{R_2^2-R_1^2} &,\quad R_1 \leq r < R_2,\\
1 &,\quad R_2 \leq r
\end{cases}
\end{equation}
and $\theta_{\rm max}=\theta(R_2)$ is the maximum kick angle given by
\begin{equation}
\theta_{\rm max} = \theta(R_2) = \frac{2LI_T(1\pm\beta_e\beta_p)}{4\pi\epsilon_0  \cdot \left(\frac{q}{p_0}\right)_p \cdot \beta_e\beta_p c^2}\cdot\frac{1}{R_2},
\end{equation}
where $L$ is the length of the e-lens, $I_T$ the total electron beam current, $\beta_{e}$ and $\beta_{p}$ the relativistic $\beta$ of electron and proton beam, $\frac{q}{p_0}=\left(B\rho\right)_p$ the magnetic rigidity for the reference particle, $c$ the speed of light and $\epsilon_0$ the vacuum permittivity.
\begin{figure}[b]
	\includegraphics[width=0.8\linewidth]{kick_hel_lhc_no_grid}% Here is how to import EPS art
	\caption{\label{fig:hel_field} Illustration of the hollow electron beam distribution (blue), the kick experienced by the proton beam (red) and the collimators (gray).}
\end{figure}
\begin{table}[t]
	\caption{\label{tab:hel_param}%
		HL-LHC hollow electron lens parameters as in~\cite{hel_cdr}.
	}
	\begin{ruledtabular}
		\begin{tabular}{lcc}
			Geometry & Value& Unit\\
			\colrule
		Length $L$    &  3 & m\\
		Desired range of scraping positions\footnote{$\sigma$ denotes here the Beam sigma and not the collimation $\sigma$ which assumes a normalized emittance of 3.5~$\mu$m.} & 4-8 &$\sigma_p$\\
		\colrule
		Magnetic fields & & \\
		\colrule
		Gun solenoid, $B_g$ & 0.2-0.4 & T\\
		Main solenoid, $B_m$ & 2-6 & T\\
		Collector solenoid, $B_c$ & 0.2-0.4 & T\\
		Compression factor ($k=\sqrt{B_m/B_g}$) & 2.2-5.5 & -\\
		\colrule
		Electron gun & & \\
		\colrule
		Peak yield $I_e$ at 10~keV & 5.0 & A\\
		Gun perveance $P$ & 5 & $\mu$perv\\
		Inner/outer cathode radius, $R_1/R_2$ & 6.75/12.7 & mm\\
		\colrule
		High voltage modulator & & \\
		\colrule
		Cathode-anode voltage & 10.0 & kV\\
		Rise time (10\%-90\%) & 200 & ns \\
		Repetition rate & 35 & kHz
		\end{tabular}
	\end{ruledtabular}
\end{table}

The distribution together with the resulting kick are illustrated in Fig.~\ref{fig:hel_field}. The $\pm$-sign represents the two cases of the electron beam traveling in the direction of the proton beam ($v_e v_p>0$) leading to ``$-$" or in the opposite direction ($v_e v_p<0$) leading to ``$+$". For hollow electron beam collimation, electron and proton beam travel in opposite directions. Assuming HL-LHC and HEL design parameters (Table~\ref{tab:hel_param}--\ref{tab:hllhc_param}) the maximum kick of the HEL is:
\begin{equation}\label{eqn:helkick}
\theta_{\rm max,B1} = 392~\rm{nrad}
\end{equation}
%\begin{eqnarray}
%\theta_{\rm max,B1} = 392~\rm{nrad},\\
%\theta_{\rm max,B2} = 341~\rm{nrad}.
%\end{eqnarray}
for an inner radius $R_1=4\sigma_p$, outer radius $R_2=7.5\sigma_p$ and a peak current of $I_e=$\SI{5.0}{A}. Similar values are obtained for Beam~2.

\subsection{Operation modes and effects on the beam core\label{sec:hel:core}}
As depicted in Fig.~\ref{fig:hel_field} and in evidence from Eq.~\ref{eq:field_1}--\ref{eq:field_2}, the field at the beam core vanishes in the ideal case. Effects on the beam core thus only arise in case of imperfections, where possible sources are the bends of the HEL, as the electron beam crosses directly the proton beam, and distortions in the electron beam profile (see Section~\ref{sec:core}). Both sources result in non-linear kicks \cite{hel_bends_stancari,hel_model_polynomial_morozov}.

In DC operation of the HEL, these non-linear kicks stay in the shadow of the non-linearities otherwise present in the machine. Tolerances on imperfections are therefore not particularly stringent and of no concern.

The picture changes significantly in case of pulsed operation of the HEL, in which case the electron gun voltage is modulated using a white or colored noise spectrum. If the electromagnetic field of the HEL does not vanish at the proton beam core, noise is transferred not only to the halo particles, as intended, but also to the beam core with naturally much smaller amplitude. Tolerances in this case rapidly become much more stringent than in the case of DC operation and studies of the effect of the HEL on the beam core are therefore focusing on this mode of operation which is also the main topic of this paper.
\begin{table}[t]
	\caption{\label{tab:hllhc_param}%
		HL-LHC design parameters at top energy \cite{hlcdr} and parameters relevant in connection with the HEL. Optics parameters at HEL are based on a position of the HEL of $-40$~m for Beam~1 (B1) and $+40$~m for Beam~2 (B2) from IP4 using HL-LHC optics V1.3 with $\beta^{*}=0.15$~m \cite{hlv13}.
	}
	\begin{ruledtabular}
		\begin{tabular}{lccc}
			Beam parameters & Value(B1) & Value(B2) & Unit\\
			\colrule
			Beam energy  $E_{p}$  &  \multicolumn{2}{c}{7} & TeV\\
			Number of bunches $n_b$ & \multicolumn{2}{c}{2748} & - \\
			Number of particles per bunch $N_b$ & \multicolumn{2}{c}{$2.2\times 10^{11}$} & -\\
			Normalized emittance $\epsilon_{N,x/y}$ & \multicolumn{2}{c}{2.5} & $\mu$m\\
			Bunch spacing & \multicolumn{2}{c}{25} & ns\\
			\colrule
			\multicolumn{4}{l}{Optics paramters at HEL (Beam~1) \footnote{As the twiss parameters at IP4 do not change during the entire squeeze, and IP4 and the HEL are only separated by a drift space, the twiss parameters stay constant also at the HEL during the entire squeeze.}} \\
			\colrule
			$\beta_{x}$ at HEL  & 197.5 & 280.6 & m\\
			$\beta_{y}$ at HEL & 211.9 & 262.6 m\\
			Dispersion $D_{x}$ at HEL & 0.0& 0.0 & m\\
			Dispersion $D_{y}$ at HEL & 0.0& 0.0 & m\\
			Proton beam size $\sigma_{p,x}$ at HEL & 0.26 & 0.31& mm \\			
			Proton beam size $\sigma_{p,y}$ at HEL & 0.27 & 0.30 &mm \\
			\multicolumn{4}{l}{scraping position}\\ \hspace{1cm}$\sigma_{p}=\max(\sigma_{p,x},\sigma_{p,y})$ & 0.27& 0.31 & mm\\
		\end{tabular}
	\end{ruledtabular}
\end{table}

Despite the fact that pulsed operation is much more dangerous than DC operation for the proton beam core, pulsed operation is still strongly under consideration. The main benefit of pulsing the HEL is to increase the halo removal rate if a fast depletion of the halo is desired. The halo diffusion rate is increased in this case as in addition to the non-linear kick, noise is exerted on the halo particles of the proton beam. For the HL-LHC two different pulsing patterns are considered:
\begin{itemize}
	\item \textbf{random excitation:} The electron gun voltage is modulated between:
	\begin{eqnarray}
		U_{\mathrm{e-gun}}&=&a\cdot U_{\mathrm{max}}\\
		& &+(1-a)\cdot \mathrm{ran}(0,1)\cdot U_{\mathrm{max}}
	\end{eqnarray}
	with $U_{\mathrm{max}}$ the equivalent voltage in DC operation, $a$ the modulation strength with $a\in[0,1]$, and $\mathrm{ran}(0,1)$ a uniformly distributed random number between~[0,1].
	\item \textbf{resonant excitation:} The HEL is switched on only every $k^{\mathrm{th}}$ turn. The excitation can then be represented by:
	\begin{eqnarray}\label{intro:eqn:1}
	f(t)&=&\sum_{p=-\infty}^{+\infty}\delta(t-n\cdot(kT)),
	\end{eqnarray}
	where $p$ is the turn number and $T$ the revolution time, and its Fourier series by:
	\begin{eqnarray}\label{intro:eqn:2}
	f(t)=\Sh_{kT}(t)
	&=&\frac{1}{kT}\sum_{n=-\infty}^{+\infty}e^{2\pi i f_nt} \\
	& & \quad \text{with} \ f_n=\frac{n}{k}f_{\rm rev},
	\end{eqnarray}
	where $\Sh_{kT}$ is the Dirac comb. As can be seen from the Fourier series, $k^{\mathrm{th}}$~turn pulsing in general drives $k^{\mathrm{th}}$ order resonances \cite{md_sim_hel_res_ex_fitterer}. Historically, the $k^{\mathrm{th}}$~turn pulsing was used in regular operation in the Tevatron for abort cleaning \cite{hel_tevatron_abortgap_zhang}.
\end{itemize}
In general, the random excitation mode is the most effective and operationally easiest pulsing mode, but also the most stringent in terms of tolerances on the residual field at the beam core as it induces white noise and thus excites all beam frequencies. As the resonant excitation acts on specific resonances, the pulsing pattern has to be chosen according to the machine and beam parameters. On the one hand, this makes the operation more difficult as one has to find the most efficient $k^{\mathrm{th}}$~turn pulsing for each operation mode, on the other hand, it also relaxes the tolerances on the residual field at the beam core as, depending on the machine and beam configuration, it might be possible to find a resonant excitation pattern, which mainly affects the halo and not the core.

\section{Sources of residual field in the proton beam core region and first order estimates\label{sec:core}}
With the current HEL layout (see Fig.~\ref{fig:hel_layout}), parasitic kicks on the proton beam core can arise in the central region (main solenoid) due to profile imperfections in the electron beam and at the entrance and exit of the HEL, where electron and proton beam overlap. Currently, no HEL is installed in the LHC, and the kick on the proton beam core must therefore be approximated to first order by a dipole kick for the experiments at the LHC. During the experiments presented in this paper, the dipole kick with the corresponding excitation pattern was applied with the LHC transverse damper (see Section~\ref{sec:adt}).

Leaping ahead of the more detailed derivations in Sec.~\ref{core:sec:1} and Sec.~\ref{core:sec:2}, the expected kick to first order from the HEL bends is estimated to be~(Eqn.~\ref{eqn:kick_bends}):
\begin{equation}
	\Delta x'_{\rm{bends}},\Delta y'_{\rm{bends}}\leq \SI{0.5}{nrad},
\end{equation}
and for the central region due to profile imperfections~(Eqn.~\ref{eqn:kick_central}):
\begin{equation}
	\Delta x'_{\rm{central \ region}}, \Delta y'_{\rm{central \ region}} \leq \SI{19.6}{nrad}
\end{equation}
where the current HEL design parameters (see Table~\ref{tab:hllhc_param}--\ref{tab:hel_param}) are used. The contribution from the central region is clearly dominating.

\subsection{Uncompensated kicks from HEL bends}
\label{core:sec:1}
As reference for the estimate of the dipole component originating from the HEL bends, the symplectic map derived in \cite{hel_bends_stancari} is used. In this case the HEL bends are modeled as a cylindrical pipe with a static charge distribution of 1~A and 5~keV electrons bend in the horizontal plane. In case of an U-shaped HEL, the transverse dipole kicks at entrance and exit add up, while for a S-shaped HEL the kicks compensated each other. For this reason a S-shape has been chosen for the HL-LHC HEL (see Fig.~\ref{fig:hel_layout}). In this case, uncompensated kicks therefore only arise from imperfections, which can originate from profile imperfections and current fluctuations. As a first rough estimate, we assume in this paper 10\% fluctuations between the entrance and exit and that the kicks from entrance and exit due to these imperfections add up. Using the electric field calculations in \cite{hel_bends_stancari}, the maximum values for the integrated electric field are then around
\begin{equation}
\int_{z_1}^{z_2} E_{x,y} dz= 10 \ \mathrm{kV}
\end{equation}
assuming an \SI{1}{A}, \SI{5}{keV} electron beam.
%For \SI{7}{TeV} protons and neglecting magnetic effects, this yields a kick of:
%\begin{equation}
%\Delta x'=\Delta y'= \SI{1.4}{nrad},
%\end{equation}
%Scaling to the HEL design parameters of a \SI{5}{A} and \SI{10}{keV} electron beam (Table~\ref{tab:hel_param}), one obtains:
%\begin{equation}
%\int_{z_1}^{z_2} E_{x} dz= 36 \ \mathrm{kV} \Rightarrow \Delta x'= \q{5.1}{nrad}
%\end{equation}
Neglecting magnetic effects and scaling to HL-LHC and HEL design parameters (Table~\ref{tab:hel_param}--\ref{tab:hllhc_param}) this yields a kick of:
\begin{equation}
\int_{z_1}^{z_2} E_{x} dz= 36 \ \mathrm{kV} \Rightarrow \Delta x'= \q{5.1}{nrad}
\end{equation}
Assuming now 10\% fluctuation between entrance and exit kick, the expected kick from the HEL bends is
\begin{equation}\label{eqn:kick_bends}
\Delta x', \Delta y'= \q{0.5}{nrad}.
\end{equation}

\subsection{Kicks in the central region (main solenoid)}
\label{core:sec:2}
For a perfectly uniform, annular and radially symmetric profile, the field in the region of the proton beam core vanishes (see Eqn.~\ref{eq:field_2} and Fig.~\ref{fig:hel_field}). Fields at the location of the proton beam core, thus only arise in case of profile imperfections of the electron beam, in particular if the radial symmetry is broken.

>>>>

In a recent review, the need for such an active halo control system for HL-LHC has been assessed with the conclusion that it would considerably increase the margin and reduce the risk for machine protection compared to the current LHC collimation system being a passive halo control system ~\cite{helreview}. In view of the need of active halo control for HL-LHC and for future high power accelerators like HE-LHC and FCC-hh~\cite{helhcparam2011,fcc_coll_ipac2017}, different active halo control methods have been studied in recent years~\cite{helreview_bruce}, of which the hollow electron lens (HEL) is considered the superior device for the HL-LHC~\cite{helreview}. The beneficial effect of a future HEL for HL-LHC in terms of machine protection would be however abrogated if the beam core was negatively affected leading ultimately to a degradation of the performance due to increased particle losses and emittance growth. In this paper, we will concentrate on this aspect. We will summarize possible sources of perturbations of the proton beam core focusing in particular on the case of pulsed operation and the beam experiments at the LHC conducted in this context.

This paper is structured as follows: Section~\ref{sec:hel} gives an introduction to the concept of HELs and summarizes the design parameters of the HL-LHC HELs. Section~\ref{sec:core} is dedicated to describing the sources of a residual field from the HEL in the core region. Section~\ref{sec:exp} then describes in detail the two beam experiments conducted at the LHC studying the effect of pulsed operation on the proton beam core.

%This includes the detailed analysis of the observed losses, emittance and transverse beam distribution changes. To the knowledge of the authors, the observed distribution changes presented in this paper have never been measured before in the context of a resonant excitation. A resonant excitation has been previously studied at the Tevatron in the context of the HEL experiments \cite{hel_tevatron_stancari} and the abort gap cleaning used in operation \cite{hel_tevatron_abortgap_zhang}. Both studies however only concentrated on the losses and emittance changes without measuring the detailed changes of the distribution itself. In addition, the presented experiments also provide scalings of the losses and emittance growth with the excitation amplitude. This allows to provide first tolerances for the residual field of the HL-LHC HEL at the beam core and also to compare the obtained results with simulations and thus benchmark the same.

\section{Hollow electron lens for HL-LHC\label{sec:hel}}
\subsection{General Overview\label{sec:hel:intro}}
Electron lenses are in general DC or pulsed low energy electron beams, where the electron beam is generated in an electron gun, then guided and confined by strong solenoids and finally dumped onto a collector. As an example the conceptual design of the HL-LHC HEL is shown in Fig.~\ref{fig:hel_layout}.
\begin{figure}[h]
	\includegraphics[width=1.0\linewidth]{hel_layout_epbeam}% Here is how to import EPS art
	\caption{\label{fig:hel_layout} Layout of HL-LHC HEL.}
\end{figure}

\begin{figure}[b]
	\includegraphics[width=0.8\linewidth]{kick_hel_lhc_no_grid}% Here is how to import EPS art
	\caption{\label{fig:hel_field} Illustration of the hollow electron beam distribution (blue), the kick experienced by the proton beam (red) and the collimators (gray).}
The circulating beam, in case of the LHC the proton beam, is then affected by the electromagnetic field of the electron beam. For the application of active halo control, the electron beam needs to generate an electromagnetic field at the location of the halo particles while leaving the core unperturbed. This field distribution can be achieved by using a uniform hollow distribution in radius $r=\sqrt{x^2+y^2}$ with inner radius $R_1$ and outer radius $R_2$. In this case, the circulating proton beam experiences a radial kick $\theta(r)$
\begin{equation}\label{eq:field_1}
\theta(r)=\frac{f(r)}{(r/R_2)}\cdot \theta_{\rm max},
\end{equation}
where $f(r)$ is a shape function with
\begin{equation}\label{eq:field_2}
f(r) =
\begin{cases} 0 &,\quad r< R_1,\\
\frac{r^2-R_1^2}{R_2^2-R_1^2} &,\quad R_1 \leq r < R_2,\\
1 &,\quad R_2 \leq r
\end{cases}
\end{equation}
and $\theta_{\rm max}=\theta(R_2)$ is the maximum kick angle given by
\begin{equation}\label{hel_kick_max}
\theta_{\rm max} = \theta(R_2) = \frac{2LI_T(1\pm\beta_e\beta_p)}{4\pi\epsilon_0  \cdot \left(\frac{q}{p_0}\right)_p \cdot \beta_e\beta_p c^2}\cdot\frac{1}{R_2},
\end{equation}
with $L$ the length of the HEL, $I_T$ the total electron beam current, $\beta_{e}$ and $\beta_{p}$ the relativistic $\beta$ of electron and proton beam, $\frac{q}{p_0}=\left(B\rho\right)_p$ the magnetic rigidity for the proton beam reference particle, $c$ the speed of light and $\epsilon_0$ the vacuum permittivity. The distribution together with the resulting kick are illustrated in Fig.~\ref{fig:hel_field}. The $\pm$-sign in Eq.~\ref{hel_kick_max} represents the two cases of the electron beam traveling in the direction of the proton beam ($v_e v_p>0$) leading to ``$-$" or in the opposite direction ($v_e v_p<0$) leading to ``$+$". For hollow electron beam collimation, the beams are chosen to counterrotate. Assuming HL-LHC and HEL design parameters (see Table~\ref{tab:hllhc_param}--\ref{tab:hel_param}) the maximum kick of the HEL is:
\begin{equation}\label{eqn:helkick}
\theta_{\rm max,B1} = 392~\rm{nrad}
\end{equation}
%\begin{eqnarray}
%\theta_{\rm max,B1} = 392~\rm{nrad},\\
%\theta_{\rm max,B2} = 341~\rm{nrad}.
%\end{eqnarray}
for an inner radius of $R_1=4\sigma_p$, outer radius $R_2=7.5\sigma_p$, peak current of $I_e=$\SI{5.0}{A} and Beam~1 of the LHC. Similar values are obtained for Beam~2.
\begin{table}[t]
	\caption{\label{tab:hllhc_param}%
		HL-LHC design parameters at top energy \cite{hlcdr} and parameters relevant in connection with the HEL. Optics parameters at HEL are based on a position of the HEL of $-40$~m for Beam~1 (B1) and $+40$~m for Beam~2 (B2) from IP4 using HL-LHC optics V1.3 with $\beta^{*}=0.15$~m \cite{hlv13}.
	}
	\begin{ruledtabular}
		\begin{tabular}{lccc}
			Beam parameters & Value(B1) & Value(B2) & Unit\\
			\colrule
			Beam energy  $E_{p}$  &  \multicolumn{2}{c}{7} & TeV\\
			Number of bunches $n_b$ & \multicolumn{2}{c}{2748} & - \\
			Number of particles per bunch $N_b$ & \multicolumn{2}{c}{$2.2\times 10^{11}$} & -\\
			Normalized emittance $\epsilon_{N,x/y}$ & \multicolumn{2}{c}{2.5} & $\mu$m\\
			Bunch spacing & \multicolumn{2}{c}{25} & ns\\
			\colrule
			\multicolumn{4}{l}{Optics paramters at HEL (Beam~1) \footnote{As the twiss parameters at IP4 do not change during the entire squeeze, and IP4 and the HEL are only separated by a drift space, the twiss parameters stay constant also at the HEL during the entire squeeze.}} \\
			\colrule
			$\beta_{x}$ at HEL  & 197.5 & 280.6 & m\\
			$\beta_{y}$ at HEL & 211.9 & 262.6 m\\
			Dispersion $D_{x}$ at HEL & 0.0& 0.0 & m\\
			Dispersion $D_{y}$ at HEL & 0.0& 0.0 & m\\
			Proton beam size $\sigma_{p,x}$ at HEL & 0.26 & 0.31& mm \\			
			Proton beam size $\sigma_{p,y}$ at HEL & 0.27 & 0.30 &mm \\
			\multicolumn{4}{l}{scraping position}\\ \hspace{1cm}$\sigma_{p}=\max(\sigma_{p,x},\sigma_{p,y})$ & 0.27& 0.31 & mm\\
		\end{tabular}
	\end{ruledtabular}
\end{table}
\begin{table}[t]
	\caption{\label{tab:hel_param}%
		HL-LHC hollow electron lens parameters as in~\cite{hel_cdr}.
	}
	\begin{ruledtabular}
		\begin{tabular}{lcc}
			Geometry & Value& Unit\\
			\colrule
			Length $L$    &  3 & m\\
			Desired range of scraping positions\footnote{$\sigma$ denotes here the Beam sigma and not the collimation $\sigma$ which assumes a normalized emittance of 3.5~$\mu$m.} & 4-8 &$\sigma_p$\\
			\colrule
			Magnetic fields & & \\
			\colrule
			Gun solenoid, $B_g$ & 0.2-0.4 & T\\
			Main solenoid, $B_m$ & 2-6 & T\\
			Collector solenoid, $B_c$ & 0.2-0.4 & T\\
			Compression factor ($k=\sqrt{B_m/B_g}$) & 2.2-5.5 & -\\
			\colrule
			Electron gun & & \\
			\colrule
			Peak yield $I_e$ at 10~keV & 5.0 & A\\
			Gun perveance $P$ & 5 & $\mu$perv\\
			Inner/outer cathode radius, $R_1/R_2$ & 6.75/12.7 & mm\\
			\colrule
			High voltage modulator & & \\
			\colrule
			Cathode-anode voltage & 10.0 & kV\\
			Rise time (10\%-90\%) & 200 & ns \\
			Repetition rate & 35 & kHz
		\end{tabular}
	\end{ruledtabular}
\end{table}

\subsection{Operation modes and effects on the beam core\label{sec:hel:core}}
As depicted in Fig.~\ref{fig:hel_field} and in evidence from Eqs.~\ref{eq:field_1}--\ref{eq:field_2}, the field at the beam core vanishes in case of an ideal HEL. Effects on the beam core thus only arise from imperfections, where possible sources are the bends of the HEL (see Section~\ref{core:sec:1}), as here the electron beam crosses directly the proton beam, and distortions in the electron beam profile (see Section~\ref{core:sec:2}). Both sources result in non-linear kicks~\cite{hel_bends_stancari,hel_model_polynomial_morozov}. In DC operation, these non-linear kicks stay in the shadow of the non-linearities otherwise present in the machine. Tolerances on imperfections are therefore not particularly stringent and of no concern. The picture changes significantly in case of pulsed operation, in which case the electron gun voltage is modulated using a white or colored noise spectrum. If the electromagnetic field of the HEL does not vanish at the proton beam core, noise is transferred not only to the halo particles, as intended, but also to the beam core. Tolerances on the residual field in this case rapidly become much more stringent than in the case of DC operation. Studies of the effect of the HEL on the beam core are therefore focusing on this mode of operation which will also be the main subject of this paper.

The main benefit of pulsed HEL operation is the increase in halo removal rate particularly relevant if a fast depletion of the halo is desired. The halo diffusion rate increases in this case as in addition to the non-linear kick, noise is exerted on the halo particles. For the HL-LHC two different pulsing patterns are considered for this purpose:
\begin{itemize}
	\item \textbf{random excitation:} The electron gun voltage is modulated between:
	\begin{eqnarray}
		U_{\mathrm{e-gun}}&=&a\cdot U_{\mathrm{max}}\\
		& &+(1-a)\cdot \mathrm{ran}(0,1)\cdot U_{\mathrm{max}}
	\end{eqnarray}
	with $U_{\mathrm{max}}$ the equivalent voltage in DC operation, $a$ the modulation strength with $a\in[0,1]$, and $\mathrm{ran}(0,1)$ a uniformly distributed random number between~[0,1].
	\item \textbf{resonant excitation:} The HEL is switched on only every $k^{\mathrm{th}}$ turn. The excitation can then be represented by:
	\begin{eqnarray}\label{intro:eqn:1}
	f(t)&=&\sum_{p=-\infty}^{+\infty}\delta(t-n\cdot(kT)),
	\end{eqnarray}
	where $p$ is the turn number and $T$ the revolution time, and its Fourier series by:
	\begin{eqnarray}\label{intro:eqn:2}
	f(t)=\Sh_{kT}(t)
	&=&\frac{1}{kT}\sum_{n=-\infty}^{+\infty}e^{2\pi i f_nt} \\
	& & \quad \text{with} \ f_n=\frac{n}{k}f_{\rm rev},
	\end{eqnarray}
	where $\Sh_{kT}$ is the Dirac comb. As can be seen from the Fourier series, $k^{\mathrm{th}}$~turn pulsing in general drives $k^{\mathrm{th}}$ order resonances \cite{md_sim_hel_res_ex_fitterer}. Historically, the $k^{\mathrm{th}}$~turn pulsing was used in regular operation in the Tevatron for abort cleaning \cite{hel_tevatron_abortgap_zhang}.
\end{itemize}
In general, the random excitation mode is the most effective and operationally easiest pulsing mode, but also the most stringent in terms of tolerances on the residual field at the proton beam center as it induces white noise exciting all beam frequencies. As the resonant excitation acts on specific resonances, the pulsing pattern has to be chosen according to the machine and beam parameters. On the one hand, this makes the operation more difficult as one has to find the most efficient $k^{\mathrm{th}}$~turn pulsing for each operation mode, on the other hand, it also relaxes the tolerances on the residual field at the beam core as, depending on the machine and beam configuration, it might be possible to find a resonant excitation pattern, which mainly affects the halo and not the beam core.

\section{Sources of residual field in the proton beam core region and first order estimates\label{sec:core}}
With the current HEL layout as shown in Fig.~\ref{fig:hel_layout}, parasitic kicks on the proton beam core can arise in the central region due to profile imperfections in the electron beam and at the entrance and exit of the HEL, where electron and proton beam overlap. Currently, no HEL is installed in the LHC, and the kick on the proton beam core must therefore be emulated with the help of other devices already installed in the LHC. For this purpose the kick is approximated to first order by a dipole kick. During the experiments presented in this paper, the dipole kick with the corresponding excitation pattern could then be applied with the LHC transverse damper as will be described in detail in Section~\ref{sec:adt}. Leaping ahead of the more detailed derivations in Sec.~\ref{core:sec:1} and \ref{core:sec:2}, the expected kick to first order from the HEL bends is estimated to be~(Eq.~\ref{eqn:kick_bends}):
\begin{equation}
	\Delta x'_{\rm{bends}},\Delta y'_{\rm{bends}}\approx \SI{0.5}{nrad},
\end{equation}
and for the central region due to profile imperfections~(Eq.~\ref{eqn:kick_central}):
\begin{equation}
	\Delta x'_{\rm{central \ region}}, \Delta y'_{\rm{central \ region}} \approx \SI{19.6}{nrad}
\end{equation}
where the current HEL design parameters are used (see Table~\ref{tab:hllhc_param}--\ref{tab:hel_param}). The contribution from the central region is clearly dominating.

\subsection{Uncompensated kicks from HEL bends}
\label{core:sec:1}
As reference for the estimate of the dipole component originating from the HEL bends, the symplectic map derived in \cite{hel_bends_stancari} is used. In this case the HEL bends are modeled as a cylindrical pipe with a static charge distribution of 1~A and 5~keV electrons deflected in the horizontal plane. In case of an U-shaped HEL, the transverse dipole kicks at entrance and exit add up, while for a S-shape the kicks compensate each other. For this reason a S-shape has been chosen for the HL-LHC HEL (see Fig.~\ref{fig:hel_layout}). Assuming and S-shaped HEL, uncompensated kicks therefore only arise from imperfections, which can originate from profile imperfections and current fluctuations. As a first estimate, we assume in this paper 10\% fluctuations between the entrance and exit and, in addition, that the kicks from entrance and exit due to these imperfections add up. Using the electric field calculations in \cite{hel_bends_stancari}, the maximum values for the integrated electric field are then
\begin{equation}
\int_{z_1}^{z_2} E_{x,y} dz= 10 \ \mathrm{kV}
\end{equation}
given an \SI{1}{A}, \SI{5}{keV} electron beam.
%For \SI{7}{TeV} protons and neglecting magnetic effects, this yields a kick of:
%\begin{equation}
%\Delta x'=\Delta y'= \SI{1.4}{nrad},
%\end{equation}
%Scaling to the HEL design parameters of a \SI{5}{A} and \SI{10}{keV} electron beam (Table~\ref{tab:hel_param}), one obtains:
%\begin{equation}
%\int_{z_1}^{z_2} E_{x} dz= 36 \ \mathrm{kV} \Rightarrow \Delta x'= \q{5.1}{nrad}
%\end{equation}
Neglecting magnetic effects and scaling to HL-LHC and HEL design parameters (Tables~\ref{tab:hllhc_param}--\ref{tab:hel_param}) yields a kick of:
\begin{equation}
\int_{z_1}^{z_2} E_{x} dz= 36 \ \mathrm{kV} \Rightarrow \Delta x'= \q{5.1}{nrad}
\end{equation}
Assuming now 10\% fluctuation between entrance and exit kick, the expected kick from the HEL bends is
\begin{equation}\label{eqn:kick_bends}
\Delta x', \Delta y'\approx \q{0.5}{nrad}.
\end{equation}

% insert table here to put it on top of page
\begin{table*}
	\caption{\label{tab:md_param}%
		Beam parameters and machine configuration as used in the two resonant excitation experiments in 2016 and 2017 \cite{resexmd2016,resexmd2017}. The plane of the excitation is abbreviated with H for horizontal, V for vertical and H+V horizontal and vertical at the same time.
	}
	\begin{ruledtabular}
		\begin{tabular}{lcc}
			Parameter & Experiment 2017 & Experiment 2016  \\
			\colrule
			beam &\multicolumn{2}{c}{Beam~1} \\
			beam energy &\multicolumn{2}{c}{injection energy, 450 GeV} \\\hline
			single bunch intensity &\multicolumn{2}{c}{$0.7\times10^{11}$} \\
			normalized emittance &\multicolumn{2}{c}{$2.5-3.5~\mathrm{\mu m}$} \\
			$4\sigma$ bunch length & \multicolumn{2}{c}{1.3~ns}\\
			$1\sigma$ bunch length & \multicolumn{2}{c}{9.7~cm}\\
			number of bunches & $3\times72=216$ bunches & $12\times4=48$ single bunches\\
			& (+ 1 pilot + 12 nominal) & \\\hline
			injection optics, $\beta^*=11$~m & standard optics 2017 & standard optics 2016\\
			Landau damping octupoles  & \multicolumn{2}{c}{$I_{\mathrm{MO}}=\pm19.6~\mathrm{A}$, explicitly +19.6 A for MOF circuit and}\\
			& \multicolumn{2}{c}{-19.6~A for MOD circuit (standard 2016 settings)}\\\hline
			working point $(Q_x,Q_y)$ & (62.27,60.295) & (64.28,59.31)\\
			chromaticity $(Q'_x,Q'_y)$ & \multicolumn{2}{c}{(+15,+15)}\\\hline
			pulsing patterns  & $7^{\mathrm{th}}$~turn H,V,H+V;& $7^{\mathrm{th}}$~turn H;\\
			& $8^{\mathrm{th}}$~turn H,V,H+V;& $10^{\mathrm{th}}$~turn V;\\
			& random  H,V,H+V;& \\
		\end{tabular}
	\end{ruledtabular}
\end{table*}
% insert table here to put it on top of page
